Bootstrapping Multi-atlas Segmentation Using Multiple Automatically Generated Templates for the Segmentation of the Whole Hippocampus and Subfields

### 2013-12-12 paper.Rnw
Updated Experiment 2 methods to address reviews

\textbf{Introduction:} 
Advances in image segmentation of magnetic resonance images (MRI) have
demonstrated that multi-atlas approaches improve segmentation accuracy and
precision over regular atlas-based approaches.  These approaches often rely on
a large number of such manually segmented atlases (e.g.  30-80) that take
significant time and expertise to produce. We present an algorithm, \mb{} (
({\textbf M}ultiple {\textbf A}utomatically {\textbf Ge}nerated {\textbf
T}emplates), for the automatic segmentation of the hippocampus that minimizes
the number of atlases needed while still achieving similar accuracy to
multi-atlas approaches. Thus, our method acts as an accurate multi-atlas
approach when using special, hard-to-define atlases that are laborious to
construct. \\
\textbf{Method:} \mb{} works by propagating atlas segmentations to a template library,
formed from a subset of target images, via transformations estimated by non-linear
image registration. The resultant segmentations are then propagated to each
target image and fused using a label fusion method. We conduct a 10-fold Monte
Carlo cross-validation of whole hippocampal segmentation on a pool of 69 ADNI
subjects over a range of parameter settings, and a leave-one-out
cross-validation (LOOCV) of hippocampal subfield segmentation using five
high-resolution atlases. Two final experiments assess \mb{} when applied to
first episode psychosis and Alzheimer's disease populations, and \mb{}
segmentations are compared with existing automated methods (\fsl{}, \freesurfer{},
\maper{}) and biases are explored.\\
\textbf{Results:} Using nine atlases and 19 template images, \mb{} achieves a
mean Dice's Similarity Coefficient (DSC) of 0.84 over 10-folds of Monte Carlo
cross-validation on 69 subjects relative to semi-automated \gs{} segmentations,
and shows significantly lower variability in DSC than multi-atlas segmentation.
In a LOOCV, \mb{} reproduces the volumes of the cornu ammonis (CA) 1; CA4/dentate
gyrus (DG); and strata radiatum (SR), strata lacunosum (SL), and strata
moleculare (SM) hippocampal subregions with a percent error in volume that is
at or lower than that produced by image resampling. \mb{} produces hippocampal
volumes in a first episode psychosis patient population that are highly
correlated with expert manual segmentation volumes (Pearson $r = 0.877, t =
16.244, p < 0.001$). Compared to \fsl{} and \freesurfer{}, \mb{} shows much smaller
fixed volume bias (within $250 mm^3$ on average) to \gs{} segmentations available
from ADNI, as well as a conservative, rather than exaggerated, proportional
volume bias.\\ \textbf{Conclusion:} We demonstrate that \mb{} produces accurate
hippocampal segmentations using only 5 atlases over different hippocampal
definitions, disease populations, and acquisition types, as well as showing
that accurate identification of the hippocampal subfields is possible.
\end{abstract}

@@

The hippocampus is a brain structure situated in the medial temporal
lobe, and has long been associated with learning and memory
\citep{DenHeijer2012,Scoville2000}. The hippocampus is of interest to
clinical neuroscientists because it is implicated in many forms of brain
dysfunction, including Alzheimer's disease \citep{Sabuncu2011} and
schizophrenia \citep{Narr2004,Karnik-Henry2012}. In neuroimaging studies,
structural magnetic resonance images (MRI) are often used for the volumetric
assessment of the hippocampus.  As such, accurate segmentation of the
hippocampus and its subregions in MRI is a necessary first step to better
understand the inter-individual variability of subject neuroanatomy.

@@

\paragraph{\adnidataset{} dataset} 
Data used in the preparation of this article were obtained from the Alzheimer’s
Disease Neuroimaging Initiative (ADNI) database (adni.loni.ucla.edu). The ADNI
was launched in 2003 by the National Institute on Aging (NIA), the National
Institute of Biomedical Imaging and Bioengineering (NIBIB), the Food and Drug
Administration (FDA), private pharmaceutical companies and non-profit
organizations, as a \$60 million, 5-year public- private partnership. The primary
goal of ADNI has been to test whether serial magnetic resonance imaging (MRI),
positron emission tomography (PET), other biological markers, and clinical and
neuropsychological assessment can be combined to measure the progression of mild
cognitive impairment (MCI) and early Alzheimer’s disease (AD).  Determination of
sensitive and specific markers of very early AD progression is intended to aid
researchers and clinicians to develop new treatments and monitor their
effectiveness, as well as lessen the time and cost of clinical trials.

The Principal Investigator of this initiative is Michael W. Weiner, MD, VA
Medical Center and University of California – San Francisco. ADNI is the result
of efforts of many co- investigators from a broad range of academic institutions
and private corporations, and subjects have been recruited from over 50 sites
across the U.S. and Canada. The initial goal of ADNI was to recruit 800 subjects
but ADNI has been followed by ADNI-GO and ADNI-2.  To date these three protocols
have recruited over 1500 adults, ages 55 to 90, to participate in the research,
consisting of cognitively normal (CN) older individuals, people with early or
late MCI, and people with early AD. The follow up duration of each group is
specified in the protocols for ADNI-1, ADNI-2 and ADNI-GO. Subjects originally
recruited for ADNI-1 and ADNI-GO had the option to be followed in ADNI-2. For
up-to-date information, see www.adni-info.org.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%                                                                     %%%%%
%%%%%        Experiment 2: Hippocampal Subfield Cross-Validation          %%%%%
%%%%%                                                                     %%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Experiment 2: Hippocampal Subfield Cross-Validation}
In this experiment, the accuracy of the \mb{} algorithm on hippocampal subregion
segmentation is tested using a leave-one-out cross-validation (LOOCV) design on
two image sets. The optimal parameter settings for \mb{} found in Experiment 1 are
used in this experiment.

\paragraph{Winterburn Atlases dataset} 
The Winterburn atlases \citep{Winterburn2013} are digital hippocampal
segmentations of five in-vivo $300\mu$ isotropic T1-weighted MR images. The
segmentations include subfield segmentations for the cornu ammonis (CA) 1; CA2
and CA3; CA4 and dentate gyrus; subiculum; and strata radiatum (SR), strata
lacunosum (SL), and strata moleculare (SM). Subjects in the Winterburn atlases
range in age from 29-57 years (mean age of 37), and include two males and three
females.

In addition to the high-resolution scans distributed as part of the Winterburn
atlases, we also obtained separate 3T T1 BRAVO acquisitions (0.9mm-isotropic
voxels) of four of the five Winterburn atlas subjects.  Images were acquired on
a 3T GE Discovery MR 750 system (General Electric, Milwaukee, WI) using an
8-channel head coil with the enhanced fast gradient recalled echo 3-dimensional
acquisition protocol, EFGRE-BRAVO, with the following parameters: $TE/TR/TI =
3.0ms/6.7ms/650ms$, flip angle $=8^\circ$, $FOV=15.8cm$, slice
thickness$=0.9mm$, 176 in-plane steps for an approximate isotropic resolution
of 0.9mm dimension voxels.

\paragraph{Experiment details} 
Leave-one-out cross-validation (LOOCV) is an approach in which the method to be
validated is given all but one item in a dataset as training data (in our case,
atlas library images and labels), and the output accuracy is measured on the
left-out item. This is done, in turn, for each item in the dataset.

@@

In part 2a of this experiment, we have modified the LOOCV experimental
design in order to compare \mb{} segmentation accuracy with the accuracy of image
resampling error. The five $0.3mm$-isotropic voxel Winterburn atlas images and
labels are used as the atlas library, and LOOCV is conducted using the five Winterburn
atlas images subsampled (using trilinear interpolation) to
0.9mm-isotropic voxel resolution (referred to as the {\em Subsampled} dataset)
as input subjects. Subsampling the target images allows us to assess \mb{} in a
typical segmentation scenario (high-resolution atlases and lower-resolution
subjects).  The template library is composed of all five subject images, plus
an additional set of 16 of healthy subjects 3T T1 images acquired separately
(Table \ref{tab:WAval-healthy-demographics}).  Images were acquired on a 3T GE
Discovery MR 750 system (General Electric, Milwaukee, WI) using an 8-channel
head coil with the enhanced fast gradient recalled echo 3-dimensional
acquisition protocol, EFGRE-BRAVO, with the following parameters: $TE/TR/TI =
3.0ms/6.7ms/650ms$, flip angle=$8^\circ$ , $FOV = 15.3cm$, slice
thickness$=0.9mm$, 170 in-plane steps for an approximate isotropic resolution
of 0.9mm dimension voxels.

The optimal size of template library, registration method, and label fusion
method found in Experiment 1 are used.  As with traditional LOOCV, each subject
image is segmented by \mb{} with that subject's image excluded from the atlas
library.  In part 2b of this experiment, we carry out a similar validation in
which the subject images are separately acquired T1 BRAVO images of four of the
five Winterburn atlas subjects (referred to as the {\em T1 BRAVO} dataset).
Figure \ref{fig:LOOCV-schematic} illustrates the experimental set up of both
parts of this experiment.

@@

\paragraph{Evaluation method}  
To assess the \mb{} LOOCV segmentations we compute the relative percent error in 
hippocampal volume with respect to the full resolution Winterburn atlas 
segmentations. In addition, by computing the relative error in volume of the
Winterburn atlas labels resampled (with nearest-neighbour interpolation) to
0.9mm-isotropic voxels, we obtain a baseline error against which to assess \mb{}
segmentations.

@@

In general, across hippocampal subregions the percent error in volume of \mb{}
segmentations (relative to the full-resolution Winterburn atlas segmentation)
compares favourably to the error resulting from image resampling (Figure
\ref{fig:WAval-vol-boxplot}). In particular, the CA1, CA4/DG, and SR/SL/SM
subregions all show a percent error in volume that is at or lower than
resampling error. The Subiculum and CA2/CA3 subregions show distinctly larger
percent error in volume than is found through resampling.

@@

% - take away: MB produces subregion volumes that are comparable or better than

>>>>

\textbf{Introduction:} 
Advances in image segmentation of magnetic resonance images (MRI) have
demonstrated that multi-atlas approaches improve segmentation accuracy and
precision over regular atlas-based approaches.  These approaches often rely on
a large number of such manually segmented atlases (e.g.  30-80) that take
significant time and expertise to produce. We present an algorithm, \mb{} (
({\textbf M}ultiple {\textbf A}utomatically {\textbf Ge}nerated {\textbf
T}emplates), for the automatic segmentation of the hippocampus that minimizes
the number of atlases needed while still achieving similar accuracy to
multi-atlas approaches. Thus, our method acts as an accurate multi-atlas
approach when using special, hard-to-define atlases that are laborious to
construct. \\
\textbf{Method:} \mb{} works by propagating atlas segmentations to a template library,
formed from a subset of target images, via transformations estimated by non-linear
image registration. The resultant segmentations are then propagated to each
target image and fused using a label fusion method. We conduct a 10-fold Monte
Carlo cross-validation of whole hippocampal segmentation on a pool of 69 ADNI
subjects over a range of parameter settings, and a leave-one-out
cross-validation (LOOCV) of hippocampal subfield segmentation using five
high-resolution atlases. Two final experiments assess \mb{} when applied to
first episode psychosis and Alzheimer's disease populations, and \mb{}
segmentations are compared with existing automated methods (\fsl{}, \freesurfer{},
\maper{}) and biases are explored.\\
\textbf{Results:} Using nine atlases and 19 template images, \mb{} achieves a
mean Dice's Similarity Coefficient (DSC) of 0.84 over 10-folds of Monte Carlo
cross-validation on 69 subjects relative to semi-automated \gs{} segmentations,
and shows significantly lower variability in DSC than multi-atlas segmentation.
In a LOOCV, \mb{} reproduces the volumes of the cornu ammonis (CA) 1; CA4/dentate
gyrus (DG); and strata radiatum (SR), strata lacunosum (SL), and strata
moleculare (SM) hippocampal subfields with a percent error in volume that is
at or lower than that produced by image resampling. \mb{} produces hippocampal
volumes in a first episode psychosis patient population that are highly
correlated with expert manual segmentation volumes (Pearson $r = 0.877, t =
16.244, p < 0.001$). Compared to \fsl{} and \freesurfer{}, \mb{} shows much smaller
fixed volume bias (within $250 mm^3$ on average) to \gs{} segmentations available
from ADNI, as well as a conservative, rather than exaggerated, proportional
volume bias.\\ \textbf{Conclusion:} We demonstrate that \mb{} produces accurate
hippocampal segmentations using only 5 atlases over different hippocampal
definitions, disease populations, and acquisition types, as well as showing
that accurate identification of the hippocampal subfields is possible.
\end{abstract}

@@

The hippocampus is a brain structure situated in the medial temporal
lobe, and has long been associated with learning and memory
\citep{DenHeijer2012,Scoville2000}. The hippocampus is of interest to
clinical neuroscientists because it is implicated in many forms of brain
dysfunction, including Alzheimer's disease \citep{Sabuncu2011} and
schizophrenia \citep{Narr2004,Karnik-Henry2012}. In neuroimaging studies,
structural magnetic resonance images (MRI) are often used for the volumetric
assessment of the hippocampus.  As such, accurate segmentation of the
hippocampus and its subfields in MRI is a necessary first step to better
understand the inter-individual variability of subject neuroanatomy.

@@

\paragraph{\adnidataset{} dataset} 
Data used in the preparation of this article were obtained from the Alzheimer's
Disease Neuroimaging Initiative (ADNI) database (adni.loni.ucla.edu). The ADNI
was launched in 2003 by the National Institute on Aging (NIA), the National
Institute of Biomedical Imaging and Bioengineering (NIBIB), the Food and Drug
Administration (FDA), private pharmaceutical companies and non-profit
organizations, as a \$60 million, 5-year public-private partnership. The primary
goal of ADNI has been to test whether serial magnetic resonance imaging (MRI),
positron emission tomography (PET), other biological markers, and clinical and
neuropsychological assessment can be combined to measure the progression of mild
cognitive impairment (MCI) and early Alzheimer's disease (AD).  Determination of
sensitive and specific markers of very early AD progression is intended to aid
researchers and clinicians to develop new treatments and monitor their
effectiveness, as well as lessen the time and cost of clinical trials.

The Principal Investigator of this initiative is Michael W. Weiner, MD, VA
Medical Center and University of California San Francisco. ADNI is the result
of efforts of many co- investigators from a broad range of academic institutions
and private corporations, and subjects have been recruited from over 50 sites
across the U.S. and Canada. The initial goal of ADNI was to recruit 800 subjects
but ADNI has been followed by ADNI-GO and ADNI-2. To date these three protocols
have recruited over 1500 adults, ages 55 to 90, to participate in the research,
consisting of cognitively normal (CN) older individuals, people with early or
late MCI, and people with early AD. The follow up duration of each group is
specified in the protocols for ADNI-1, ADNI-2 and ADNI-GO. Subjects originally
recruited for ADNI-1 and ADNI-GO had the option to be followed in ADNI-2. For
up-to-date information, see www.adni-info.org.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%                                                                     %%%%%
%%%%%        Experiment 2: Hippocampal Subfield Cross-Validation          %%%%%
%%%%%                                                                     %%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Experiment 2: Hippocampal Subfield Cross-Validation}
The previous experiment assesses \mb{} peformance on whole hippocampus
segmentation. In this experiment, we assess \mb{} hippocampal subfield
segmentation of standard 3T T1-weighted images using a leave-one-out
cross-validation (LOOCV) design.

\paragraph{Winterburn Atlases} 
The Winterburn atlases \citep{Winterburn2013} are digital hippocampal
segmentations of five in-vivo $0.3mm$-isotropic T1-weighted MR images. The
segmentations include subfield segmentations for the cornu ammonis (CA) 1; CA2
and CA3; CA4 and dentate gyrus; subiculum; and strata radiatum (SR), strata
lacunosum (SL), and strata moleculare (SM). Subjects in the Winterburn atlases
range in age from 29-57 years (mean age of 37), and include two males and three
females.

\paragraph{Experiment details} 
Leave-one-out cross-validation (LOOCV) is a validation approach in which an
algorithm is given all but one item in a dataset as training data (in our case,
atlas images and labels) and then is applied to the left-out item. This is done,
in turn, for each item in the dataset and the output across all items is
evaluated.

In this experiment, the high-resolution $0.3mm^3$ voxel Winterburn atlases are
used as the \mb{} atlas library, but in each round of LOOCV a $0.9mm^3$ voxel
target image corresponding to the left-out atlas is used so as to evaluate \mb{}
on standard 3T T1-weighted resolution images.

Specifically, two sets of $0.9mm^3$ voxel target images are used in this
experiment: separately acquired T1 acquisitions of four of the five Winterburn
atlas subjects (referred to as the {\em BRAVO} set because of the scan
parameters used, described below), and computationally subsampled versions of
the Winterburn atlas images (the {\em Subsampled} set).

The 3T T1 BRAVO acquisitions of four of the five Winterburn atlas subjects were
separately obtained within a short time of the original atlas image
acquisitions.  Images were acquired on a 3T GE Discovery MR 750 system (General
Electric, Milwaukee, WI) using an 8-channel head coil with the enhanced fast
gradient recalled echo 3-dimensional acquisition protocol, EFGRE-BRAVO, with the
following parameters: $TE/TR/TI = 3.0ms/6.7ms/650ms$, flip angle $=8^\circ$,
$FOV=15.8cm$, slice thickness$=0.9mm$, 176 in-plane steps for an approximate
isotropic resolution of 0.9mm dimension voxels.

Image subsampling of the Winterburn atlases was performed using trilinear
interpolation techniques.

The template library is composed of all $0.9mm^3$ voxel target images
(either {\em BRAVO} or {\em Subsampled}), plus an additional set of 16 of
healthy subjects 3T T1 images acquired separately (Table
\ref{tab:WAval-healthy-demographics}). These images were acquired on a 3T GE
Discovery MR 750 system (General Electric, Milwaukee, WI) using an 8-channel
head coil with the enhanced fast gradient recalled echo 3-dimensional
acquisition protocol, EFGRE-BRAVO, with the following parameters: $TE/TR/TI =
3.0ms/6.7ms/650ms$, flip angle=$8^\circ$ , $FOV = 15.3cm$, slice
thickness$=0.9mm$, 170 in-plane steps for an approximate isotropic resolution of
0.9mm dimension voxels.

The optimal size of template library, registration method, and label fusion
method found in Experiment 1 are used. Figure \ref{fig:LOOCV-schematic}
illustrates schematically the experimental set up.

@@

@@

\paragraph{Evaluation method}  
In this experiment we must evaluate \mb{} on a task which is ill-defined since
there are no segmentation protocols of the hippocampal subfields for T1 images
at $0.9mm^3$ voxels (in fact, whether subfields can even be differentiated at
this resolution is debated \citep{who?}). Therefore, we evaluate the {\it
precision} with \mb{} produces hippocampal subfields that correspond to the
segmentation protocol used by the given atlas library images.

Specifically, we compute the relative percent error in volume of the \mb{}-produced
hippocampal subfields with respect to the full-resolution Winterburn atlas
segmentations. In addtion, by directly resampling the Winterburn atlas
segmentations to $0.9mm^3$ voxel images (using nearest-neighbour interpolation)
we can obtain a baseline relative percent error to compare \mb{} against.

@@

In general, across hippocampal subfields the percent error in volume of \mb{}
segmentations (relative to the full-resolution Winterburn atlas segmentation)
compares favourably to the error resulting from image resampling (Figure
\ref{fig:WAval-vol-boxplot}). In particular, the CA1, CA4/DG, and SR/SL/SM
subfields all show a percent error in volume that is at or lower than
resampling error. The Subiculum and CA2/CA3 subfields show distinctly larger
percent error in volume than is found through resampling.

@@

% - take away: MB produces subfields volumes that are comparable or better than
